\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2021-10-13 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd
last review of paper

---
author:
  - name: Sahir Rai Bhatnagar*
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon*
    affiliation: 'University of Manitoba'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James A. Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title: "\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
abstract: |
  | In epidemiological studies of time-to-event data, a quantity of interest to the clinician is their patient's risk of an event. However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the  estimating function. The baseline hazard then needs to be estimated separately using a non-parametric approach. This leads to step-wise estimates of the survival function that are difficult to interpret. Using case-base sampling, Hanley & Miettinen (2009) explained how parametric hazard functions can be estimated using logistic regression. Their approach naturally leads to estimates of the survival function that are smooth-in-time.
  | In this paper, we present the \CRANpkg{casebase} R package, a comprehensive and flexible toolkit for parametric survival analysis. We describe how the case-base framework can be used in more complex settings: non-linear functions of time and non-proportional hazards, competing risks, and variable selection. Our package also includes an extensive array of visualization tools to complement the analysis. We illustrate all these features through three different case studies.
  | * SRB and MT contributed equally to this work.
preamble: >
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{makecell}
  \usepackage{tabu}
output: rticles::rjournal_article
bibliography: bhatnagar-turgeon-islam-hanley-saarela.bib 
fig_width: 12
fig_height: 9
fig_caption: true
editor_options: 
  chunk_output_type: console
---

@@

The semiparametric Cox model has become the default approach to survival analysis 
even though Cox himself later suggested he would prefer to model the hazard function 
directly. In a 1994 interview with Professor Nancy Reid, Sir David Cox was asked 
how he would model a set of censored survival data, to which he responded: "I 
think I would normally want to tackle problems parametrically ... and if you want 
to do things like predict the outcome for a particular patient, it's much more 
convenient to do that parametrically" [@reid1994conversation]. Indeed, the most 
relevant quantity in a clinical setting is often the 5- or 10-year risk of 
experiencing a certain event given the patient's particular circumstances. 
Unfortunately, the most reported metric from a Cox model is the (potentially 
time-dependent) hazard ratio (HR), which ignores the duration of follow-up or 
is subject to selection bias [@hernan2010hazards]. The covariate adjusted 
survival curve overcomes these limitations, and is arguably a more important 
summary measure to report than the HR. While stepwise survival curves can be 
computed with the Cox model, they require a second step to separately estimate 
the baseline hazard [@breslow1972discussion].

Several authors have since pursued fully parametric approaches that made the 
fitting of smooth hazard functions much more transparent and intuitive through 
generalized linear models. The key feature of these procedures is splitting the 
time axis into discrete intervals. [@whitehead1980fitting] showed the 
equivalence between a Cox model and a Poisson regression with a parameter for 
each event time. [@copenhagen2012needs] provides a nice exposition of this 
equivalence with a real data example and supporting R code for computing 
standard errors. [@arjas1987logistic] and [@efron1988logistic] treated each 
patient-day as a Bernoulli random variable with probability equal to the 
discrete hazard rate. A potential issue with these approaches is that the 
number of time bins need to be chosen by the data analyst. A fine grouping 
of times may result in few (or none) events per interval, which then leads 
to instability in the Newton-Raphson procedure for estimation 
[@kalbfleisch2011statistical]. A coarse grouping could potentially mask 
nonlinear trends in the hazard function.

Rather than discretizing time, Hanley & Miettinen [-@hanley2009fitting] 
selected a discrete set of person-time coordinates (`person-moments') in 
continuous time from all observed follow-up experience constituting the 
study base to obtain the logistic form likelihood expression for the hazard 
function. More specifically, all person-moments when the event of interest 
occurred are selected as the case series, complemented by a randomly sampled 
base series of person-moments serving as controls. This approach allows 
flexible modeling of the hazard function by including time as a covariate 
(e.g. using splines or general additive models). Furthermore, time-dependent 
covariates can be modeled through interactions with time. In short, Hanley & 
Miettinen [-@hanley2009fitting] use the well-understood logistic regression 
for directly modeling the hazard function, but do not require an arbitrary 
choice of splitting time.

In this article, we present the \CRANpkg{casebase} R package [@casebase-package] 
which implements the Hanley & Miettinen [-@hanley2009fitting] approach for 
fitting fully parametric hazard models and covariate adjusted survival curves 
using the familiar interface of the \code{glm} function. Our implementation 
allows for straightforward extensions to other models such as penalized 
regression for variable selection and competing-risk analysis. In addition, 
we provide functions for exploratory data analysis and visualizing the estimated 
quantities such as the hazard function, survival curve and its standard errors. 
The ultimate goal of our package is to make fitting flexible hazards accessible 
to more endusers with the hopes that they will favor reporting absolute risks 
over hazard ratios.

In what follows, we first recall some theoretical details on case-base sampling 
and its use for estimating parametric hazard functions. We then give a short 
review of existing R packages that implement comparable features as \pkg{casebase}. 
Next, we provide some details about the implementation of case-base sampling in 
our package, and we give a brief survey of its main functions. This is followed 
by three case studies that illustrate the flexibility and capabilities of 
\pkg{casebase}. We show how the same framework can be used for non-linear 
functions of time and non-proportional hazards, competing risks and variable 
selection via penalized regression. Finally, we end the article with a 
discussion of the results and of future directions.

@@

As discussed in Hanley & Miettinen [-@hanley2009fitting], the key idea behind case-base sampling is to consider the entire study base as an infinite collection of *person moments*. These person moments are indexed by both an individual in the study and a time point, and therefore each person moment has a covariate profile, an exposure status, and an outcome status attached to it. By comparing person moments at which an event occurred with person moments at which no event occurred, we can extract information about hazards and survival.

@@

To complement the case series, we sample person moments for the base series. To do so, we model the base series sampling mechanism using non-homogeneous Poisson processes $R_i(t) \in \{0, 1, 2, \ldots\}$; the person-moments where $dR_i(t) = 1$ constitute the base series. The process $Q_{i}(t) = R_i(t) + N_{i}(t)$ then counts both the case and base series person-moments contributed by individual $i$. The processes $R_i(t)$ are typically defined by the user via its intensity function $\rho_i(t)$. The process $Q_{i}(t)$ is characterized by $E[dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)dt + \rho_i(t)dt$.

If the hazard function $\lambda_{i}(t; \theta)$ is parametrized in terms of $\theta$, we could define an estimator $\hat{\theta}$ by maximization of the likelihood expression
$$L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^{\min(t_i,\tau)} \lambda_i(t; \theta) dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{dN_{i}(t)},$$
where $\prod_{t\in[0,u)}$ represents a product integral from $0$ to $u$, and where $t_i$ is the event time for individual $i$. However, the integral over time makes the computation and maximization of $L_0(\theta)$ challenging.

@@

\begin{equation}
\log(\lambda(t; X)) = \beta_0 + \beta_1 X. \label{eq:exp}
\end{equation}

In this model, the instantaneous failure rate is independent of $t$.\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study. This is also known as the \dfn{memoryless property} \citep{kalbfleisch2011statistical}.}

@@

Case-base sampling can also be used in the context of competing-risk analyses. Assuming there are $J$ competing events, we can show that each person-moment's contribution to the likelihood is of the form

$$\frac{\lambda_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^J\lambda_j(t)},$$

where $N_j(t)$ is the counting process associated with the event of type $j$ and $\lambda_j(t)$ is the corresponding cause-specific hazard function. As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1.\end{align*}}

@@

At the time of writing, a cursory examination of CRAN's \ctv{Survival} Task View reveals that there are over 250 packages related to survival analysis [@survTaskView]. For the purposes of this article, we restricted our review to packages that implement at least one of the following features: parametric modeling, non-proportional hazard models, competing risk analysis, penalized estimation, and CIF estimation. By searching for appropriate keywords in the \code{DESCRIPTION} file of these packages, we found 60 relevant packages. These 60 packages were then manually examined to determine which ones are comparable to \pkg{casebase}. In particular, we excluded packages that were focused on a different set of problems, such as frailty and multi-state models. The remaining 14 packages appear in Table \ref{tab:surv-pkgs}, along with some of the functionalities they offer.

Parametric survival models are implemented in a handful of packages, each differing in the underlying data distribution: \CRANpkg{CFC} [-@mahani2015bayesian], \CRANpkg{flexsurv} [-@flexsurv], \CRANpkg{SmoothHazard} [-@smoothHazard], \CRANpkg{rstpm2} [-@clements_liu], \CRANpkg{mets} [-@scheike2014estimating], and \CRANpkg{survival} [-@survival-package]. For example, \pkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas both \pkg{flexsurv} and \pkg{survival} allow users to supply any distribution of their choice. \pkg{flexsurv}, \pkg{SmoothHazard}, \pkg{mets} and \pkg{rstpm2} can model the effect of time using splines, which allows flexible modeling of the hazard function. As discussed above, \pkg{casebase} can model any parametric family whose log-hazard can be expressed as a linear combination of covariates (including time). Therefore, our package is more general in that it allows the user to model any linear or non-linear transformation of time including splines and higher order polynomials. Also, by including interaction terms between covariates and time, it also allows users to fit (non-proportional) time-varying coefficient models. However, unlike \pkg{flexsurv}, we do not explicitly model any shape parameter.

@@

Next, several R packages implement methodologies for competing risk analysis; for a different perspective on this topic, see Mahani & Sharabiani [-@mahani2015bayesian]. The package \pkg{survival} provides functionality for competing-risk analysis and multistate modelling. The package \CRANpkg{cmprsk} provides methods for cause-specific subdistribution hazards, such as in the Fine-Gray model [-@fine1999proportional]. On the other hand, the package \pkg{CFC} estimates cause-specific CIFs from unadjusted, non-parametric survival functions. Our package \pkg{casebase} also provides functionalities for competing risk analysis by estimating parametrically the cause-specific hazards. From these quantities, we can then estimate the cause-specific CIFs.

Finally, several packages include functions to estimate the survival function and the CIF. The corresponding methods generally fall into two categories: transformation of the estimated hazard function, and semi-parametric estimation of the baseline hazard. The first category broadly corresponds to parametric survival models, where the full hazard is explicitly modeled. Using this estimate, the survival function and the CIF can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CDF} below). Packages providing this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and \pkg{survival}. Our package \pkg{casebase} also follows this approach for both single-event and competing-risk analyses. The second category outlined above broadly corresponds to semi-parametric models. These models do not model the full hazard function, and therefore the baseline hazard needs to be estimated separately in order to estimate the survival function. This is achieved using semi-parametric estimators (e.g. Breslow's estimator) or parametric estimators (e.g. spline functions). Packages that implement this approach include \pkg{riskRegression}, \pkg{rstpm2}, \pkg{survival}, and \pkg{glmnet}. As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the survival function, whereas the second category often produces estimates in the form of stepwise functions.

@@

\caption{Comparison of various R packages for survival analysis based on several defining features. \textbf{Competing Risks}: whether or not an implementation for competing risks is present. \textbf{Allows Non PH}: permits models for non-proportional hazards. \textbf{Penalized Regression}: allows for a penalty term on the regression coefficients when estimating hazards (e.g. lasso or ridge). \textbf{Splines}: permits a flexible fit on time through the use of splines. \textbf{Parametric}: implementation for parametric models. \textbf{Semi-parametric}: implementation for semi-parametric models. \textbf{Interval/left censoring}: models for interval and left-censoring. If this is not selected, the package only handles right-censoring. \textbf{Risk estimates}: computation for survival curves, cumulative incidence or cumulative hazard is readily available.}

@@

Population-time plots are a descriptive visualization of incidence density, where aggregate person-time is represented by area and events as points within the area. The case-base sampling approach described above can be visualized in the form of a population time plot. These plots are extremely informative graphical displays of survival data and should be one of the first steps in an exploratory data analysis. The \code{popTime} function and \code{plot} method facilitate this task:

@@

```{r erspc-data, eval = eval_cs1, echo = c(-1:-3)}
data("ERSPC")
ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
                       levels = c(0,1), 
                       labels = c("Control group", "Screening group"))

pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
                               event = "DeadOfPrCa", exposure = "ScrArm")
inherits(pt_object, "popTime")
attr(pt_object, "exposure")
```

@@

```{r plot-stratified-erspc-data, eval = eval_cs1, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plots for both treatment arms in the ERSPC dataset. The gray area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely thin rectangles (person-moments). More events are observed at later follow-up times, motivating the use of non-constant hazard models."}
label_number_si <- function (accuracy = 1, unit = NULL, sep = NULL, ...) {
    sep <- if (is.null(unit)) 
        ""
    else " "
    scales:::force_all(accuracy, ...)
    function(x) {
        breaks <- c(0, 10^c(K = 3, M = 6, B = 9, T = 12))
        n_suffix <- cut(abs(x), breaks = c(unname(breaks), Inf), 
            labels = c(names(breaks)), right = FALSE)
        n_suffix[is.na(n_suffix)] <- ""
        suffix <- paste0(sep, n_suffix, unit)
        scale <- 1/breaks[n_suffix]
        scale[which(scale %in% c(Inf, NA))] <- 1
        number(x, accuracy = accuracy, scale = unname(scale), 
            suffix = suffix, ...)
    }
}
plot(pt_object, 
     add.base.series = TRUE,
     ratio = 0.7,
     facet.params = list(ncol = 2),
     ribbon.params = list(fill = "gray50"),
     case.params = list(size = 0.8),
     base.params = list(size = 0.8)) + 
  paper_gg_theme +  
  scale_y_continuous(labels = label_number_si()) + 
  ylab("Population Size")
```

>>>>

---
author:
  - name: Sahir Rai Bhatnagar*
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon*
    affiliation: 'University of Manitoba'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James A. Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title: "\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
abstract: |
  | In epidemiological studies of time-to-event data, a quantity of interest to the clinician is their patient's risk of an event. However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the estimating function. As a consequence, the focus has been on reporting hazard ratios instead of survival curves. Indeed, reporting patient risk requires a separate estimation of the baseline hazard. Using case-base sampling, Hanley & Miettinen (2009) explained how parametric hazard functions can be estimated in continuous-time using logistic regression. Their approach naturally leads to estimates of the survival function that are smooth-in-time.
  | In this paper, we present the \CRANpkg{casebase} R package, a comprehensive and flexible toolkit for parametric survival analysis. We describe how the case-base framework can be used in more complex settings: non-linear functions of time and non-proportional hazards, competing risks, and variable selection. Our package also includes an extensive array of visualization tools to complement the analysis. We illustrate all these features through three different case studies.
  | * SRB and MT contributed equally to this work.
preamble: >
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{makecell}
  \usepackage{tabu}
output: rticles::rjournal_article
bibliography: bhatnagar-turgeon-islam-hanley-saarela.bib 
fig_width: 12
fig_height: 9
fig_caption: true
editor_options: 
  chunk_output_type: console
---

@@

The semiparametric Cox model has become the default approach to survival analysis 
even though Cox himself later suggested he would prefer to model the hazard function 
directly. In a 1994 interview with Professor Nancy Reid, Sir David Cox was asked 
how he would model a set of censored survival data, to which he responded: "I 
think I would normally want to tackle problems parametrically ... and if you want 
to do things like predict the outcome for a particular patient, it's much more 
convenient to do that parametrically" [@reid1994conversation]. Indeed, the most 
relevant quantity in a clinical setting is often the 5- or 10-year risk of 
experiencing a certain event given the patient's particular circumstances. 
Unfortunately, the most reported metric from a Cox model is the (potentially 
time-dependent) hazard ratio (HR), which ignores the duration of follow-up and 
is subject to selection bias [@hernan2010hazards]. The covariate-adjusted 
survival curve overcomes these limitations, and it is arguably a more important 
summary measure to report than the HR. While stepwise survival curves can be 
computed with the Cox model, they require a second step to separately estimate 
the baseline hazard [@breslow1972discussion].

Several authors have since pursued fully parametric approaches that made the 
fitting of smooth survival curves much more transparent and intuitive through 
generalized linear models. The key feature of these procedures is splitting the 
time axis into discrete intervals. Whitehead [-@whitehead1980fitting] showed the 
equivalence between a Cox model and a Poisson regression with a parameter for 
each event time; Carstensen [-@copenhagen2012needs] provides a nice exposition of this 
equivalence with a real data example and supporting R code for computing 
standard errors. Arjas & Haara [-@arjas1987logistic] and Efron [-@efron1988logistic] treated each 
patient-day as a Bernoulli random variable with probability equal to the 
discrete hazard rate. A potential issue with these approaches is that the 
number of time bins need to be chosen by the data analyst. On the one hand, a fine grouping 
of times may result in few (or none) events per interval, which then leads 
to instability in the Newton-Raphson procedure for estimation 
[@kalbfleisch2011statistical]. On the other hand, a coarse grouping could potentially mask 
nonlinear trends in the hazard function.

Rather than discretizing time, Hanley & Miettinen [-@hanley2009fitting] 
selected a discrete set of person-time coordinates (`person-moments') in 
continuous time from all observed follow-up experience constituting the 
study base. By doing so, they obtained a likelihood expression for the hazard 
function that is equivalent to that of logistic regression. More specifically, all person-moments when the event of interest 
occurred are selected as the case series, complemented by a randomly sampled 
base series of person-moments serving as controls. This approach allows 
flexible modeling of the hazard function by including time as a covariate 
(e.g. using splines or general additive models). Furthermore, time-dependent 
covariates can be modeled through interactions with time. In short, Hanley & 
Miettinen [-@hanley2009fitting] use the well-understood logistic regression 
for directly modeling the hazard function, without requiring a discrete-time model.

In this article, we present the \CRANpkg{casebase} R package [@casebase-package] 
which implements the Hanley & Miettinen [-@hanley2009fitting] approach for 
fitting fully parametric hazard models and covariate-adjusted survival curves 
using the familiar interface of the \code{glm} function. Our implementation 
allows for straightforward extensions to other models such as penalized 
regression for variable selection and competing-risk analysis. In addition, 
we provide functions for exploratory data analysis and visualizing the estimated 
quantities such as the hazard function, survival curve, and their standard errors. 
The ultimate goal of our package is to make fitting flexible hazards accessible 
to more end users with the hope that they will favor reporting absolute risks 
over hazard ratios.

In what follows, we first recall some theoretical details on case-base sampling 
and its use for estimating parametric hazard functions. We then give a short 
review of existing R packages that implement comparable features as \pkg{casebase}. 
Next, we provide some details about the implementation of case-base sampling in 
our package, and we give a brief survey of its main functions. This is followed 
by three case studies that illustrate the flexibility and capabilities of 
\pkg{casebase}. We show how the same framework can be used for non-linear 
functions of time and non-proportional hazards, competing risks, and variable 
selection via penalized regression. Finally, we end the article with a 
discussion of the results and of future directions.

@@

As discussed in Hanley & Miettinen [-@hanley2009fitting], the key idea behind case-base sampling is to consider the entire study base as an infinite collection of *person moments*. These person moments are indexed by both an individual in the study and a time point, and therefore each person moment has a covariate profile, an exposure status, and an outcome status (i.e. whether the event happened) attached to it. By comparing person moments at which an event occurred with person moments at which no event occurred, we can extract information about hazards and survival.

@@

To complement the case series, we sample person moments for the base series. To do so, we model the base series sampling mechanism using non-homogeneous Poisson processes $R_i(t) \in \{0, 1, 2, \ldots\}$; the person-moments where $dR_i(t) = 1$ constitute the base series. We note that the same individual can contribute multiple person-moments to the base series. The process $Q_{i}(t) = R_i(t) + N_{i}(t)$ then counts both the case and base series person-moments contributed by individual $i$. As mentioned above, the processes $R_i(t)$ are typically defined by the user via its intensity function $\rho_i(t)$. The process $Q_{i}(t)$ is characterized by $E[dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)dt + \rho_i(t)dt$.

If the hazard function $\lambda_{i}(t; \theta)$ is parametrized in terms of $\theta$, we can define an estimator $\hat{\theta}$ by maximization of the likelihood expression
$$L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^{\min(t_i,\tau)} \lambda_i(t; \theta) dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{dN_{i}(t)},$$
where $\prod_{t\in[0,u)}$ represents a product integral from $0$ to $u$, and where $t_i$ is the event time for individual $i$. However, the integral over time makes the computation and maximization of $L_0(\theta)$ challenging.

@@

\begin{equation}
\log(\lambda(t; X)) = \beta_0 + \beta_1 X. \label{eq:exp}
\end{equation}
In this model, the instantaneous failure rate is independent of $t$.\footnote{The conditional chance of failure in a time interval of specified length is the same regardless of how long the individual has been in the study. This is also known as the \dfn{memoryless property} \citep{kalbfleisch2011statistical}.}

@@

Case-base sampling can also be used in the context of competing-risk analysis. Assuming there are $J$ competing events, we can show that each person-moment's contribution to the likelihood is of the form

$$\frac{\lambda_j(t)^{dN_j(t)}}{\rho(t) + \sum_{j=1}^J\lambda_j(t)},$$
where $N_j(t)$ is the counting process associated with the event of type $j$ and $\lambda_j(t)$ is the corresponding cause-specific hazard function. As may be expected, this functional form is similar to the terms appearing in the likelihood function for multinomial regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1.\end{align*}}

@@

At the time of writing, a cursory examination of CRAN's \ctv{Survival} Task View reveals that there are over 250 packages related to survival analysis [@survTaskView]. For the purposes of this article, we restricted our review to packages that implement at least one of the following features: parametric modeling, non-proportional hazard models, competing risk analysis, penalized estimation, and Cumulative Incidence (CI) estimation. By searching for appropriate keywords in the \code{DESCRIPTION} file of these packages, we found 60 relevant packages. These 60 packages were then manually examined to determine which ones are comparable to \pkg{casebase}. In particular, we excluded packages that were focused on a different set of problems, such as frailty and multistate models. The remaining 14 packages appear in Table \ref{tab:surv-pkgs}, along with some of the functionalities they offer.

Parametric survival models are implemented in several packages, each differing in the parametric distributions available: \CRANpkg{CFC} [-@mahani2015bayesian], \CRANpkg{flexsurv} [-@flexsurv], \CRANpkg{SmoothHazard} [-@smoothHazard], \CRANpkg{rstpm2} [-@clements_liu], \CRANpkg{mets} [-@scheike2014estimating], and \CRANpkg{survival} [-@survival-package]. For example, \pkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas both \pkg{flexsurv} and \pkg{survival} allow users to supply any distribution of their choice. \pkg{flexsurv}, \pkg{SmoothHazard}, \pkg{mets} and \pkg{rstpm2} can model the effect of time using splines, which allows flexible modeling of the hazard function. As discussed above, \pkg{casebase} can model any parametric family whose log-hazard can be expressed as a linear combination of covariates (including time). Therefore, our package is more general in that it allows the user to model any linear or non-linear transformation of time including splines and higher order polynomials. Also, by including interaction terms between covariates and time, it also allows users to fit (non-proportional) time-varying coefficient models. However, unlike \pkg{flexsurv}, we do not explicitly model any shape parameter.

@@

Next, several R packages implement methodologies for competing risk analysis; for a different perspective on this topic, see Mahani & Sharabiani [-@mahani2015bayesian]. The package \pkg{survival} provides functionality for competing-risk analysis and multistate modelling. The package \CRANpkg{cmprsk} provides methods for cause-specific subdistribution hazards, such as in the Fine-Gray model [-@fine1999proportional]. On the other hand, the package \pkg{CFC} estimates cause-specific CIs from unadjusted, non-parametric survival functions. Our package \pkg{casebase} also provides functionalities for competing risk analysis by estimating parametrically the cause-specific hazards. From these quantities, we can then estimate the cause-specific CIs.

Finally, several packages include functions to estimate the survival function and the CI. The corresponding methods generally fall into two categories: transformation of the estimated hazard function, and semi-parametric estimation of the baseline hazard. The first category broadly corresponds to parametric survival models, where the full hazard is explicitly modeled. Using this estimate, the survival function and the CI can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CDF} below). Packages providing this functionality include \pkg{CFC}, \pkg{flexsurv}, \pkg{mets}, and \pkg{survival}. Our package \pkg{casebase} also follows this approach for both single-event and competing-risk analyses. The second category outlined above broadly corresponds to semi-parametric models. These models do not model the full hazard function, and therefore the baseline hazard needs to be estimated separately in order to estimate the survival function. This is achieved using semi-parametric estimators (e.g. Breslow's estimator) or parametric estimators (e.g. spline functions). Packages that implement this approach include \pkg{riskRegression}, \pkg{rstpm2}, \pkg{survival}, and \pkg{glmnet}. As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the survival function, whereas the second category often produces estimates in the form of stepwise functions.

@@

\caption{Comparison of various R packages for survival analysis. \textbf{Competing Risks}: whether an implementation for competing risks is present. \textbf{Allows Non PH}: includes models for non-proportional hazards. \textbf{Penalized Regression}: allows for a penalty term on the regression coefficients when estimating hazards (e.g. lasso or ridge). \textbf{Splines}: allows a flexible fit on time through the use of splines. \textbf{Parametric}: implementation for parametric models. \textbf{Semi-parametric}: implementation for semi-parametric models. \textbf{Interval/left censoring}: models for interval and left-censoring. If this is not selected, the package only handles right-censoring. \textbf{Risk estimates}: estimation of survival curve and cumulative incidence is available.}

@@

Population-time plots are a descriptive visualization of incidence density, where the study base is represented by area and events by points within the area. The case-base sampling approach described above can be visualized in the form of a population time plot. These plots are informative graphical displays of survival data and should be one of the first steps in an exploratory data analysis. The \code{popTime} function and \code{plot} method facilitate this task:

@@

```{r erspc-data-mutate, eval = eval_cs1, echo = FALSE}
data("ERSPC")
ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
                       levels = c(0,1), 
                       labels = c("Control group", "Screening group"))
```

```{r erspc-data, eval = eval_cs1}
pt_object <- casebase::popTime(ERSPC, time = "Follow.Up.Time",
                               event = "DeadOfPrCa", exposure = "ScrArm")
inherits(pt_object, "popTime")
attr(pt_object, "exposure")
```

```{r plot-erspc-data, eval = eval_cs1, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plot for the ERSPC dataset. \\textbf{A}: The gray area can be thought of as N = 159,893 infinitely thin horizontal rectangles ordered by length of follow-up. \\textbf{B}: The red points correspond to when death has occurred for any one of those infinitely thin rectangles. \\textbf{C}: To improve visibility, these red points are randomly redistributed along their respective x-coordinates, providing a visualization of incidence density. More events are observed at later follow-up times, motivating the use of non-constant hazard models. \\textbf{D}: The base series, a representative sample of the entire grey area, is represented by the green points."}
# For increased visibility, remove some cases
toRemove <- sample(which(ERSPC$DeadOfPrCa == 1), 
                   round(length(which(ERSPC$DeadOfPrCa == 1))*0.7))
# Create new dataset for illustration
ERSPC_sub <- ERSPC
ERSPC_sub$DeadOfPrCa[toRemove] <- 0

a <- casebase::popTime(ERSPC_sub, time = "Follow.Up.Time",
                       event = "DeadOfPrCa")

plota <- emptyPlot <- plot(a, add.case.series = FALSE, 
                           legend = FALSE) + 
  scale_y_continuous(labels = label_number_si())

plotb <- emptyPlot + geom_point(data = dplyr::filter(a, event == 1),
                                aes(x = time, y = ycoord, 
                                    color = "Case series",
                                    fill = "Case series"), 
                                show.legend = FALSE, size = 1.5, 
                                alpha = 0.5, shape = 21) + 
  theme(legend.position = "none") + 
  scale_y_continuous(labels = label_number_si())

plotc <- plot(a, add.case.series = TRUE, legend = FALSE) +
  scale_y_continuous(labels = label_number_si())

#TEMP MAINLY NEED FOR LEGEND
plotd <- plot(a, add.base.series = TRUE) +
  scale_y_continuous(labels = label_number_si())

legend_b <- get_legend(
  plotd + 
    guides(color = guide_legend(nrow = 1)) +
     theme(legend.direction = "horizontal",
           legend.justification = "center",
           legend.box.just = "bottom")
)

#remove legend
plotd <- plotd + theme(legend.position = "none")
plotspop <- cowplot::plot_grid(plota, plotb, plotc, plotd,
                               labels = c("A", "B", "C", "D"))
g <- plot_grid(plotspop, legend_b, ncol = 1, rel_heights = c(1, .1))
g
```

@@

```{r plot-stratified-erspc-data, eval = FALSE, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plots for both treatment arms in the ERSPC dataset. The gray area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely thin rectangles (person-moments). More events are observed at later follow-up times, motivating the use of non-constant hazard models."}
label_number_si <- function (accuracy = 1, unit = NULL, sep = NULL, ...) {
    sep <- if (is.null(unit)) 
        ""
    else " "
    scales:::force_all(accuracy, ...)
    function(x) {
        breaks <- c(0, 10^c(K = 3, M = 6, B = 9, T = 12))
        n_suffix <- cut(abs(x), breaks = c(unname(breaks), Inf), 
            labels = c(names(breaks)), right = FALSE)
        n_suffix[is.na(n_suffix)] <- ""
        suffix <- paste0(sep, n_suffix, unit)
        scale <- 1/breaks[n_suffix]
        scale[which(scale %in% c(Inf, NA))] <- 1
        number(x, accuracy = accuracy, scale = unname(scale), 
            suffix = suffix, ...)
    }
}
plot(pt_object, 
     add.base.series = TRUE,
     ratio = 0.7,
     facet.params = list(ncol = 2),
     ribbon.params = list(fill = "gray50"),
     case.params = list(size = 0.8),
     base.params = list(size = 0.8)) + 
  paper_gg_theme +  
  scale_y_continuous(labels = label_number_si()) + 
  ylab("Population Size")
```

Figure \ref{fig:plot-erspc-data} depicts the process of creating a population-time plot. It is built sequentially by first adding a layer for the area representing the population time in gray (Figure \ref{fig:plot-erspc-data} A), with subjects having the least amount of observation time plotted at the top of the y-axis. We immediately notice a distinctive *stepwise shape* in the population time area. This is due to the randomization of the Finnish cohorts which were carried out on January 1 of each of year from 1996 to 1999. Coupled with the uniform December 31 2006 censoring date, this led to large numbers of men with exactly 11, 10, 9 or 8 years of follow-up. Tracked backwards in time (i.e. from right to left), the population-time plot shows the recruitment pattern from its beginning in 1991, and the January 1 entries in successive years. Tracked forwards in time (i.e. from left to right), the plot for the first three years shows attrition due entirely to death (mainly from other causes). Since the Swedish and Belgian centres were the last to complete recruitment in December 2003, the minimum potential follow-up is three years. Tracked further forwards in time (i.e. after year 3) the attrition is a combination of deaths and staggered entries. As we can see, population-time plots summarise a wealth of information about the study into a simple graph.
